GSE158422: DNA methylation in Lung squamous cell carcinoma - RNA expression and integration

### 2023-11-28 figs/GSE158422_rna_int.Rmd
RNA integration first draft

CORES=12

@@

res <- mitch_calc(rna, genesets=gs_symbols, priority="effect",cores=32)

@@

# HERENOW

# all gene
dmagg <- agg(dm=dm,cores=16)
str(dmagg)
head(dmagg$gme_res_df)
dim(dm)
dim(dmagg$gme_res_df)
mm <- dmagg$gme_res_df[,"mean",drop=FALSE]
mmres <- mitch_calc(mm, genesets=gs_symbols, priority="effect",cores=32)
mmres_up <- subset(mmres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
mmres_dn <- subset(mmres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(mmres_up)
nrow(mmres_dn)
head(mmres_up,20)
head(mmres_dn,20)

# promoter
dmpr <- dm[grep("Promoter",dm$UCSC_RefGene_Group),]
dmpragg <- agg(dm=dmpr,cores=16)
str(dmpragg)
head(dmpragg$gme_res_df)
dim(dmpr)
dim(dmpragg$gme_res_df)
mmpr <- dmpragg$gme_res_df[,"mean",drop=FALSE]
mmprres <- mitch_calc(mmpr, genesets=gs_symbols, priority="effect",cores=32)
mmprres_up <- subset(mmprres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
mmprres_dn <- subset(mmprres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(mmprres_up)
nrow(mmprres_dn)
head(mmprres_up,20)
head(mmprres_dn,20)

# ex-promoter
dmx <- dm[grep("Promoter",dm$UCSC_RefGene_Group,invert=TRUE),]
dmxagg <- agg(dm=dmx,cores=16)
str(dmxagg)
head(dmxagg$gme_res_df)
dim(dmx)
dim(dmxagg$gme_res_df)
mmx <- dmxagg$gme_res_df[,"mean",drop=FALSE]
mmxres <- mitch_calc(mmx, genesets=gs_symbols, priority="effect",cores=32)
mmxres_up <- subset(mmxres$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
mmxres_dn <- subset(mmxres$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(mmxres_up)
nrow(mmxres_dn)
head(mmxres_up,20)
head(mmxres_dn,20)

# combined methylation
mmm <- merge(mmpr,mmx,by=0)
rownames(mmm) <- mmm$Row.names
mmm$Row.names = NULL
colnames(mmm) <- c("pr","x")
mmmres <- mitch_calc(mmm, genesets=gs_symbols, priority="effect",cores=32)
mmmres_sig <- subset(mmmres$enrichment_result, p.adjustMANOVA < 0.05)
head(mmmres_sig,20)
table(sign(mmmres_sig$s.pr) == sign(mmmres_sig$s.x))

@@

# three way
mmrna <- merge(mmm,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("pr","x","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.x))
table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
mtop <- head(mmrnares_sig,30)
mtop <- mtop[,c("set","s.pr","s.x","s.rna")]
rownames(mtop) <- gsub("REACTOME_","",mtop$set)
rownames(mtop) <- gsub("_"," ",rownames(mtop))
mtop$set=NULL
colnames(mtop) <- c("promoter","non-promoter","RNA")
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)

# rna vs promoter
mmrna <- merge(mmpr,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("pr","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.pr) == sign(mmrnares_sig$s.rna))
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.pr)
summary(mylm)
mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.pr)
summary(mylm2)
plot(mmrnares$enrichment_result$s.pr,mmrnares$enrichment_result$s.rna,xlab="promoter",ylab="RNA",main="S dist",col="gray",pch=19)
abline(v=0,h=0,lty=2,lwd=2)
points(mmrnares_sig$s.pr,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
abline(mylm,col="blue",lty=3,lwd=3)
abline(mylm2,col="red",lty=3,lwd=3)

# rna vs expromoter
mmrna <- merge(mmx,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("x","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=32)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.x) == sign(mmrnares_sig$s.rna))
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
mylm <- lm(mmrnares$enrichment_result$s.rna ~ mmrnares$enrichment_result$s.x)
summary(mylm)
mylm2 <- lm(mmrnares_sig$s.rna ~ mmrnares_sig$s.x)
summary(mylm2)
plot(mmrnares$enrichment_result$s.x,mmrnares$enrichment_result$s.rna,xlab="ex-promoter",ylab="RNA",main="S dist",col="gray",pch=19)
abline(v=0,h=0,lty=2,lwd=2)
points(mmrnares_sig$s.x,mmrnares_sig$s.rna,col="black",pch=19,cex=1.01)
abline(mylm,col="blue",lty=3,lwd=3)
abline(mylm2,col="red",lty=3,lwd=3)

@@

# all meth
sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsadn) <- paste(rownames(gsadn),"DN")
gsig_up <- subset(gsaup,FDR<0.05)
gsig_dn <- subset(gsadn,FDR<0.05)
gsig <- rbind(gsig_up,gsig_dn)
nrow(gsig)

# all meth
sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez, array.type="EPIC")
rownames(gsadn) <- paste(rownames(gsadn),"DN")
gsig_up <- subset(gsaup,FDR<0.05)
gsig_dn <- subset(gsadn,FDR<0.05)
gsig <- rbind(gsig_up,gsig_dn) 
nrow(gsig)

>>>>

CORES=16

@@

res <- mitch_calc(rna, genesets=gs_symbols, priority="effect",cores=16)

@@

# import for promoters
dmp <- dm[which(rownames(dm) %in%  gp$probe),]
mp <- mitch_import(x=dmp,DEtype="limma",geneTable=gp)
dim(dmp)
dim(mp)

# import bodies
dmb <- dm[which(rownames(dm) %in%  gb$probe),]
mb <- mitch_import(x=dmb,DEtype="limma",geneTable=gb)
dim(dmb)
dim(mb)

# promoter
dmp_res <- mitch_calc(mp, genesets=gs_symbols, priority="effect",cores=16)
dmp_res_up <- subset(dmp_res$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
dmp_res_dn <- subset(dmp_res$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(dmp_res_up)
nrow(dmp_res_dn)
head(dmp_res_up,20)
head(dmp_res_dn,20)

# body
dmb_res <- mitch_calc(mb, genesets=gs_symbols, priority="effect",cores=16)
dmb_res_up <- subset(dmb_res$enrichment_result,s.dist>0 & p.adjustANOVA < 0.05)
dmb_res_dn <- subset(dmb_res$enrichment_result,s.dist<0 & p.adjustANOVA < 0.05)
nrow(dmb_res_up)
nrow(dmb_res_dn)
head(dmb_res_up,20)
head(dmb_res_dn,20)

# combined methylation
mm <- merge(mp,mb,by=0)
rownames(mm) <- mm$Row.names
mm$Row.names=NULL
colnames(mm) <- c("prom","body")
mm_res <- mitch_calc(mm, genesets=gs_symbols, priority="effect",cores=16)
mm_res_sig <- subset(mm_res$enrichment_result,p.adjustMANOVA < 0.01)
nrow(mm_res_sig)
head(mm_res_sig,20)
table(sign(mm_res_sig$s.prom) == sign(mm_res_sig$s.body) )
mm_res_sig[sign(mm_res_sig$s.prom) != sign(mm_res_sig$s.body),]

@@

# three way
mmrna <- merge(mm,rna,by=0)
rownames(mmrna) <- mmrna$Row.names
mmrna$Row.names = NULL
colnames(mmrna) <- c("prom","body","rna")
mmrnares <- mitch_calc(mmrna, genesets=gs_symbols, priority="effect",cores=16)
mmrnares_sig <- subset(mmrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mmrnares_sig,20)
table(sign(mmrnares_sig$s.prom) == sign(mmrnares_sig$s.body))
table(sign(mmrnares_sig$s.prom) == sign(mmrnares_sig$s.rna))
table(sign(mmrnares_sig$s.body) == sign(mmrnares_sig$s.rna))
mtop <- head(mmrnares_sig,30)
mtop <- mtop[,c("set","s.prom","s.body","s.rna")]
rownames(mtop) <- gsub("REACTOME_","",mtop$set)
rownames(mtop) <- gsub("_"," ",rownames(mtop))
mtop$set=NULL
colnames(mtop) <- c("promoter","body","RNA")
colfunc <- colorRampPalette(c("blue", "white", "red"))
heatmap.2( as.matrix(mtop) , trace="none",margins = c(6,25), col=colfunc(25), cexRow=.6, cexCol=1)

# rna vs promoter
mprna <- merge(mp,rna,by=0)
rownames(mprna) <- mprna$Row.names
mprna$Row.names = NULL
colnames(mprna) <- c("prom","rna")
mprnares <- mitch_calc(mprna, genesets=gs_symbols, priority="effect",cores=16)
mprnares_sig <- subset(mprnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mprnares_sig,20)
table(sign(mprnares_sig$s.prom) == sign(mmrnares_sig$s.rna))
mprnares_sig <- subset(mprnares$enrichment_result, p.adjustMANOVA < 0.05)
mylm <- lm(mprnares$enrichment_result$s.rna ~ mprnares$enrichment_result$s.prom)
summary(mylm)
mylm2 <- lm(mprnares_sig$s.rna ~ mprnares_sig$s.pr)
summary(mylm2)
plot(mprnares$enrichment_result$s.prom,mprnares$enrichment_result$s.rna,xlab="promoter",
  ylab="RNA", main="S dist",col="gray",pch=19)
abline(v=0,h=0,lty=2,lwd=2)
points(mprnares_sig$s.pr,mprnares_sig$s.rna,col="black",pch=19,cex=1.01)
abline(mylm,col="blue",lty=3,lwd=3)
abline(mylm2,col="red",lty=3,lwd=3)

# rna vs body
mbrna <- merge(mb,rna,by=0)
rownames(mbrna) <- mbrna$Row.names
mbrna$Row.names = NULL
colnames(mbrna) <- c("body","rna")
mbrnares <- mitch_calc(mbrna, genesets=gs_symbols, priority="effect",cores=16)
mbrnares_sig <- subset(mbrnares$enrichment_result, p.adjustMANOVA < 0.05)
head(mbrnares_sig,20)
table(sign(mbrnares_sig$s.body) == sign(mbrnares_sig$s.rna))
mbrnares_sig <- subset(mbrnares$enrichment_result, p.adjustMANOVA < 0.05)
mylm <- lm(mbrnares$enrichment_result$s.rna ~ mbrnares$enrichment_result$s.body)
summary(mylm)
mylm2 <- lm(mbrnares_sig$s.rna ~ mbrnares_sig$s.body)
summary(mylm2)
plot(mbrnares$enrichment_result$s.body,mbrnares$enrichment_result$s.rna,xlab="ex-promoter",
  ylab="RNA",main="S dist",col="gray",pch=19)
abline(v=0,h=0,lty=2,lwd=2)
points(mbrnares_sig$s.body,mbrnares_sig$s.rna,col="black",pch=19,cex=1.01)
abline(mylm,col="blue",lty=3,lwd=3)
abline(mylm2,col="red",lty=3,lwd=3)

@@

# body
sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez,
  array.type="EPIC", genomic.features = "Body")
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez,
  array.type="EPIC", genomic.features = "Body")
rownames(gsadn) <- paste(rownames(gsadn),"DN")
gsig_up <- subset(gsaup,FDR<0.05)
gsig_dn <- subset(gsadn,FDR<0.05)
gsig <- rbind(gsig_up,gsig_dn)
nrow(gsig)
gsig_body <- gsig
gsig_body_up <- gsig_up
gsig_body_dn <- gsig_dn

# promoter
TSS = c("TSS200", "TSS1500")
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez,
  array.type="EPIC", genomic.features = TSS)
rownames(gsaup) <- paste(rownames(gsaup),"UP")
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez,
  array.type="EPIC", genomic.features = TSS)
rownames(gsadn) <- paste(rownames(gsadn),"DN")
gsig_up <- subset(gsaup,FDR<0.05)
gsig_dn <- subset(gsadn,FDR<0.05)
gsig <- rbind(gsig_up,gsig_dn)
nrow(gsig)
gsig_promoter <- gsig
gsig_promoter_up <- gsig_up
gsig_promoter_dn <- gsig_dn

Now make a Euler plot for GSA meth-RNA comparison.

```{r,compare_rna_gsa}

rna_up <- subset(msig,s.dist>0)$set
rna_dn <- subset(msig,s.dist<0)$set

gsig_body_up <- rownames(gsig_body_up)
gsig_body_dn <- rownames(gsig_body_dn)

gsig_promoter_up <- rownames(gsig_promoter_up)
gsig_promoter_dn <- rownames(gsig_promoter_dn)

vg <- list("RNA up"=rna_up,"RNA dn"=rna_dn,
  "GSA body up"=gsig_body_up, "GSA body dn"=gsig_body_dn,
  "GSA promoter up"=gsig_promoter_up, "GSA prom dn"=gsig_promoter_dn)

plot(euler(vg),quantities=TRUE)

Now make a Euler plot for LAM-RNA comparison.

```{r,compare_rna_mitch}

dmp_res_up <- dmp_res_up$set
dmp_res_dn <- dmp_res_dn$set

dmb_res_up <- dmb_res_up$set
dmb_res_dn <- dmb_res_dn$set

vm <- list("RNA up"=rna_up,"RNA dn"=rna_dn,
  "LAM body up"=dmb_res_up, "LAM body dn"=dmb_res_dn,
  "LAM promoter up"=dmp_res_up, "LAM prom dn"=dmp_res_dn)

plot(euler(vm),quantities=TRUE)
